How topos theory can help commutative algebra

### 2018-07-09 slides-como2018.tex
typo (many found by Matthias Hutzler, thanks!)

\begin{frame}{Motivating testcases}
  Let $A$ be a ring (commutative, with unit, $1 = 0$ allowed). \\
  Assume that $A$ is reduced: If $x^n = 0$, then $x = 0$.

@@

  We'll learn how the internal language of a certain well-chosen topos provides
  a way to perform the reduction in an entirely constructive manner. If so
  desired, the resulting topos-theoretic proofs can be unwinded to yield fully
  explicit, topos-free, direct proofs.

@@

  \jnote{3}{We can slightly reduce the requirements of the proof of the first
  statement by employing not a maximal ideal, but a prime ideal. The existence
  of maximal ideals is nontrivial rings is equivalent to the axiom of choice,
  while the existence of prime ideals is equivalent to the weaker Boolean Prime
  Ideal Theorem. However, this improvement doesn't change the main point; the
  proof is still wildly unconstructive.}

@@

  \jnote{1}{The internal language of a topos allows to construct objects and
  morphisms, formulate statements about them and prove such statements in a
  naive element-based language. From the internal point of view, objects look
  like sets [more precisely, types]; morphisms look like maps; epimorphisms
  look like surjections; group objects look like groups; and so on.

@@

  \jnote{2}{Any theorem which has an intuitionistic proof holds in
  the internal universe of any topos. The restriction to intuitionistic logic
  is not due to philosophical concerns; it is a fact of life that only very few
  toposes validate the law of excluded middle (for instance, sheaf toposes over
  discrete topological space do if the law of excluded middle is available in
  the metatheory). Luckily, vast amounts of mathematics can be developed in a
  purely intuitionistic setting.

@@

  \jnote{1}{Many interesting sheaves have few global sections, which is why a
  definition should as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point. More precisely,
  changing the definitions like this would yield the internal language of the
  topos of presheaves on~$X$.

  Here is an explicit example of the translation procedure. Let~$\alpha : F
  \to G$ be a morphism of sheaves on~$X$. Then (the corner quotes ``$\speak{\ldots}$''
  indicate that translation into formal language is left to the reader):
  \footnotesize
  \begin{align*}
    & X \models \speak{$\alpha$ is injective} \\[0.2em]
    \Longleftrightarrow\
    & X \models \forall s\?F\_ \forall t\?F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        U \models \forall t\?F\_ \alpha(s_0) = \alpha(t) \Rightarrow s_0 = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          V \models \alpha(s_0) = \alpha(t_0) \Rightarrow s_0 = t_0 \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          \text{for all open $W \subseteq V$: $\alpha_V(s_0|_W) = \alpha_V(t_0|_W)$ implies $s_0|_W = t_0|_W$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
        $\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}}
\end{frame}

@@

  One can also consider the classifying local of prime \emph{ideals} of~$A$.
  Its associated topological space of points is the the set of prime ideal
  of~$A$ equipped with the \emph{constructible topology}.

@@

  As an aside, the little Zariski topos is seldomly Boolean (validates the law of
  excluded middle). A necessary condition is that~$A$ is of dimension~$\leq 0$.}

@@

  No. Just as the translation from internal statements to external statements
  is entirely mechanical, so is the translation from internal proofs to
  external proofs. Any proof employing the internal language can be unwinded to
  yield an external proof not referencing the internal language.

  However, depending on the logical complexity of the statements occuring in a
  given proof, the resulting external proof might be (much) more complex than
  the internal proof. This is particularly the case if the proof involves
  double negation, for much the same reason as that in computer science,
  continuations can twist the control flow in nontrivial ways which are
  sometimes hard to understand. It is in these cases where we can extract the
  most value of the internal language, unlocking notions and proofs which might
  otherwise be hard to obtain.

@@

\begin{frame}{Revisiting the testcases}
  Let $A$ be a reduced ring.

  \jnote{1}{This slide delivers on the promise made earlier: Using the internal
  language of the little Zariski topos, we can reduce to the case of fields
  without having to employ maximal ideals, prime ideals or minimal prime ideals.
  Since the internal language machinery is itself constructive, the displayed
  proofs can be unwinded to yield external constructive proofs which don't
  reference topos theory. Details on these external proofs can soon be found in
  a forthcoming paper titled
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-wlog.pdf}{Without
  loss of generality, any reduced ring is a field}.}

  \jnote{2}{The testcase of Grothendieck's generic freeness lemma illustrates
  that the internal language of toposes can help commutative algebra even if
  one is not interested in constructivity issues. The previously known proofs
  are somewhat long and somewhat convoluted; the new proof is arguably short
  and simple.

@@

  This argument relies on the law of excluded middle and can therefore not put
  to work in the internal universe as is. However, intuitionistically we do
  have the weaker statement~$\neg\neg(\varphi \vee \neg\varphi)$, and we can
  use this law in place of the law of excluded middle to intuitionistially
  deduce that any finitely generated vector space does \emph{not not} have a
  basis.}

@@

\begin{frame}{Properties of the affine line}
  \begin{itemize}
    \item $\affl$ is a local ring:
    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
    \text{$y$ \inv} \]
    \item $\affl$ is a field:
    \begin{align*}
      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
    \end{align*}
    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
    \varepsilon^2 = 0 \}$.
    \item Any map $\affl \to \affl$ is a polynomial.
    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
    does \emph{not not} have a zero.
  \end{itemize}

@@

  \jnote{3}{An anlogue of synthetic quasicoherence holds in the classifying
  topos of rings, demonstrating that even presheaf toposes can validate
  interesting nontrivial nongeometric sequents.

@@

  \jnote{1}{Toposes, and also more specifically classifying toposes, originated
  in algebraic geometry. It is therefore deeply embarassing that as of now,
  still very little is known about the theories classified by the major toposes
  in active use by algebraic geometers.
  \begin{center}
    \includegraphics[width=0.99\textwidth]{wraith-generic-galois-theory} \\
    \footnotesize
    Gavin Wraith. Generic Galois theory of local rings. 1979.
  \end{center}}

>>>>

\begin{frame}{Motivating test cases}
  Let $A$ be a ring (commutative, with unit, $1 = 0$ allowed). \\
  Assume that $A$ is reduced: If $x^n = 0$, then $x = 0$.

@@

  We'll learn how the internal language of a certain well-chosen topos provides
  a way to perform the reduction in an entirely constructive manner. If so
  desired, the resulting topos-theoretic proofs can be unwound to yield fully
  explicit, topos-free, direct proofs.

@@

  \jnote{3}{We can slightly reduce the requirements of the proof of the first
  statement by employing not a maximal ideal, but a prime ideal. The existence
  of maximal ideals in nontrivial rings is equivalent to the axiom of choice,
  while the existence of prime ideals is equivalent to the weaker Boolean Prime
  Ideal Theorem. However, this improvement doesn't change the main point; the
  proof is still wildly nonconstructive.}

@@

  \jnote{1}{The internal language of a topos allows to construct objects and
  morphisms of the topos, formulate statements about them and prove such
  statements \emph{in a naive element-based language}. From the internal point
  of view, objects look like sets [more precisely, types]; morphisms look like
  maps; monomorphisms look like injections; epimorphisms
  look like surjections; group objects look like groups; and so on.

@@

  \jnote{2}{Any theorem which has an intuitionistic proof holds in
  the internal universe of any topos. The restriction to intuitionistic logic
  is not due to philosophical concerns; it is a fact of life that only very few
  toposes validate the law of excluded middle (for instance, sheaf toposes over
  discrete topological spaces do if the law of excluded middle is available in
  the metatheory). Luckily, vast amounts of mathematics can be developed in a
  purely intuitionistic setting.

@@

  \jnote{1}{Many interesting sheaves have few global sections, which is why a
  definition such as ``$U \models \forall s \? F\_ \varphi(s)$ iff $U \models
  \varphi(s_0)$ for all~$s_0 \in F(U)$'' would miss the point.

  Here is an explicit example of the translation procedure. Let~$\alpha : F
  \to G$ be a morphism of sheaves on~$X$. Then (the corner quotes ``$\speak{\ldots}$''
  indicate that translation into formal language is left to the reader):
  \footnotesize
  \begin{align*}
    & X \models \speak{$\alpha$ is injective} \\[0.2em]
    \Longleftrightarrow\
    & X \models \forall s\?F\_ \forall t\?F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        U \models \forall t\?F\_ \alpha(s_0) = \alpha(t) \Rightarrow s_0 = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          V \models \alpha(s_0) = \alpha(t_0) \Rightarrow s_0 = t_0 \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          \text{for all open $W \subseteq V$: $\alpha_V(s_0|_W) = \alpha_V(t_0|_W)$ implies $s_0|_W = t_0|_W$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
        $\alpha_U(s) = \alpha_U(t)$ implies $s = t$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}}
\end{frame}

@@

  One can also consider the classifying locale of prime \emph{ideals} of~$A$.
  Its associated topological space of points is the set of prime ideals
  of~$A$ equipped with the \emph{constructible topology}.

@@

  As an aside, the little Zariski topos is rarely Boolean (validates the law of
  excluded middle). A necessary condition is that~$A$ is of dimension~$\leq 0$.}

@@

  No. Just as the translation from internal statements to external statements
  is entirely mechanical, so is the translation from internal proofs to
  external proofs. Any proof employing the internal language can be unwound to
  yield an external proof not referencing the internal language.

  However, depending on the logical complexity of the statements occurring in a
  given proof, the resulting external proof might be (much) more complex than
  the internal proof. This is particularly the case if the proof involves
  double negation, for much the same reason as that in computer science,
  continuations can twist the control flow in nontrivial ways which are
  sometimes hard to understand. It is in these cases where we can extract the
  most value of the internal language, unlocking notions and proofs which might
  otherwise be hard to obtain.

@@

\begin{frame}{Revisiting the test cases}
  Let $A$ be a reduced ring.

  \jnote{1}{This slide delivers on the promise made earlier: Using the internal
  language of the little Zariski topos, we can reduce to the case of fields
  without having to employ maximal ideals, prime ideals or minimal prime ideals.
  Since the internal language machinery is itself constructive, the displayed
  proofs can be unwound to yield external constructive proofs which don't
  reference topos theory. Details on these external proofs can soon be found in
  a forthcoming paper titled
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-wlog.pdf}{Without
  loss of generality, any reduced ring is a field}.}

  \jnote{2}{The test case of Grothendieck's generic freeness lemma illustrates
  that the internal language of toposes can help commutative algebra even if
  one is not interested in constructivity issues. The previously known proofs
  are somewhat long and somewhat convoluted; the new proof is arguably short
  and simple.

@@

  This argument relies on the law of excluded middle and can therefore not put
  to work in the internal universe as is. However, intuitionistically we do
  have the weaker statement~$\neg\neg(\varphi \vee \neg\varphi)$, and we can
  use this law in place of the law of excluded middle to intuitionistically
  deduce that any finitely generated vector space does \emph{not not} have a
  basis.}

@@

\begin{frame}{Properties of the affine line}
  \begin{itemize}
%   \item $\affl$ is a local ring:
%   \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
%   \text{$y$ \inv} \]
    \item $\affl$ is a field:
    \begin{align*}
      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
    \end{align*}
    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
    \varepsilon^2 = 0 \}$.
    \bigskip
    \item Any map $\affl \to \affl$ is a polynomial function.
    \bigskip
    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
    does \emph{not not} have a zero.
  \end{itemize}

@@

  \jnote{3}{An analogue of synthetic quasicoherence holds in the classifying
  topos of rings, demonstrating that even presheaf toposes can validate
  interesting nontrivial nongeometric sequents.

@@

  \jnote{1}{Toposes, and also more specifically classifying toposes, originated
  in algebraic geometry. It is therefore deeply embarrassing that as of now,
  still very little is known about the theories classified by the major toposes
  in active use by algebraic geometers.
  \begin{center}
    \includegraphics[width=0.99\textwidth]{wraith-generic-galois-theory} \\
    \footnotesize
    Gavin Wraith. Generic Galois theory of local rings. 1979.
  \end{center}}
